Efficient simulation of individual-based population models

### 2024-03-15 article.qmd
update

---
title: "Efficient simulation of individual-based population models"
subtitle: "The `R` package `IBMPopSim`"
author:
  - name: Daphné Giorgi 
    email: daphne.giorgi@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~giorgi/
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
  - name: Vincent Lemaire 
    corresponding: true
    email: vincent.lemaire@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~vlemaire/
    orcid: 0000-0002-0433-7722
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
  - name: Sarah Kaakai
    corresponding: true
    email: sarah.kaakai@univ-lemans.fr
    url: https://perso.univ-lemans.fr/~skaakai/
    affiliations:
      - name: Le Mans Université, CNRS, Institut du Risque et de l'Assurance 
        department: Laboratoire Manceau de Mathématiques
        url: https://lmm.univ-lemans.fr/fr/
date: last-modified
date-modified: last-modified
description: |
    TODO
abstract: >+
    The `R` Package `IBMPopSim` aims to simulate the random evolution of heterogeneous populations using stochastic Individual-Based Models (IBMs). The package enables users to simulate population evolution, in which individuals are characterized by their age and some characteristics, and the population is modified by different types of events, including births/arrivals, death/exit events, or changes of characteristics. The frequency at which an event can occur to an individual can depend on their age and characteristics, but also on the characteristics of other individuals (interactions). Such models have a wide range of applications in fields including  actuarial science, biology, ecology or epidemiology. `IBMPopSim` overcomes the limitations of time-consuming IBMs simulations by implementing new efficient algorithms  based on thinning methods, which are compiled using the `Rcpp` package while providing a user-friendly interface.
keywords: [Individual-based models, stochastic simulation, population dynamics, Poisson measures, thinning method, actuarial science, insurance portfolio simulation]
citation:
  type: article-journal
  container-title: "Computo"
  doi: "xxxx"
  url: https://computo.sfds.asso.fr/template-computo-quarto
  publisher: "Société Française de Statistique"
  issn: "2824-7795"
bibliography: references.bib
github-user: computorg
repo: "ibmpopsim_computo"
draft: true # set to false once the build is running
published: false # will be set to true once accepted
format:
  computo-html: default
  computo-pdf: default
execute:
  echo: true
---

@@

## Population {#sec-population}
## Events {#sec-events}
## Events intensity {#sec-event-intensity}
## Pathwise representation
## Thinning of Poisson measure {#sec-thinning}

>>>>

---
title: "Efficient simulation of individual-based population models"
subtitle: "The `R` package `IBMPopSim`"
author:
  - name: Daphné Giorgi 
    email: daphne.giorgi@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~giorgi/
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
  - name: Vincent Lemaire 
    corresponding: true
    email: vincent.lemaire@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~vlemaire/
    orcid: 0000-0002-0433-7722
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
  - name: Sarah Kaakai
    corresponding: true
    email: sarah.kaakai@univ-lemans.fr
    url: https://perso.univ-lemans.fr/~skaakai/
    affiliations:
      - name: Le Mans Université, CNRS, Institut du Risque et de l'Assurance 
        department: Laboratoire Manceau de Mathématiques
        url: https://lmm.univ-lemans.fr/fr/
date: last-modified
date-modified: last-modified
description: |
    TODO
abstract: >+
    The `R` Package `IBMPopSim` aims to simulate the random evolution of heterogeneous populations using stochastic Individual-Based Models (IBMs). The package enables users to simulate population evolution, in which individuals are characterized by their age and some characteristics, and the population is modified by different types of events, including births/arrivals, death/exit events, or changes of characteristics. The frequency at which an event can occur to an individual can depend on their age and characteristics, but also on the characteristics of other individuals (interactions). Such models have a wide range of applications in fields including  actuarial science, biology, ecology or epidemiology. `IBMPopSim` overcomes the limitations of time-consuming IBMs simulations by implementing new efficient algorithms  based on thinning methods, which are compiled using the `Rcpp` package while providing a user-friendly interface.
keywords: [Individual-based models, stochastic simulation, population dynamics, Poisson measures, thinning method, actuarial science, insurance portfolio simulation]
citation:
  type: article-journal
  container-title: "Computo"
  doi: "xxxx"
  url: https://computo.sfds.asso.fr/template-computo-quarto
  publisher: "Société Française de Statistique"
  issn: "2824-7795"
bibliography: references.bib
github-user: VincentLemaire
repo: "ibmpopsim_computo"
draft: true # set to false once the build is running
published: false # will be set to true once accepted
format:
  computo-html: default
  computo-pdf: default
execute:
    eval: false
    echo: true
---

@@

## Population {#sec-population}

### Notations

In the remainder of the paper, the filtered probability space is denoted by $(\Omega,\{\mathcal{F}_t \},{\mathbb{P}})$, under the usual assumptions. All processes are assumed to be càdlàg and adapted to the filtration $\{\mathcal{F}_t \}$ (for instance the history of the population) on a time interval $[0,T]$. For a càdlàg process $X$, we denote $X_{t^-} := \lim_{\genfrac{}{}{0pt}{2}{s\to t}{s<t}} X_s$.

### Individuals

An individual is represented by a triplet $I = (\tau^b, \tau^d, x) \in \mathcal{I}= {\mathbb{R}}\times \bar {\mathbb{R}}\times {\mathcal{X}}$ with:

- $\tau^b \in {\mathbb{R}}$ the date of birth,
- $\tau^d \in \bar {\mathbb{R}}$ the death date, with $\tau^d = \infty$ if the individual is still alive,
- a collection $x \in {\mathcal{X}}$ of characteristics where ${\mathcal{X}}$ is the space of characteristics.

Note that in IBMs, individuals are usually characterized by their age $a(t) =t-\tau^b$ instead of their date of birth $\tau^b$. However, using the latter is actually easier for the simulation, as it remains constant over time.

### Population process

The population at a given time $t$ is a random set
$$
    Z_t=\{ I_k \in \mathcal{I}; \; k= 1,\dots, N_t\},
$$
composed of all individuals (alive or dead) who have lived in the population before time $t$. As a random set, $Z_t$ can be represented by a random counting measure on $\mathcal{I}$ , that is an integer-valued measure $Z: \Omega \times \mathcal{I}\to \bar {\mathbb{N}}$ where for $A \in \mathcal{I}$, $Z(A)$ is the (random) number of individuals $I$ in the subset $A$. With this representation: 
$$
\begin{aligned}
    Z_t (\mathrm{d}\tau^b, \mathrm{d}\tau^d , \mathrm{d}x) &= \sum_{k=1}^{N_t} \delta_{I_k} (\tau^b, \tau^d,x), \nonumber \\
    \text{ with }  \int_{\mathcal{I}} f(\tau^b, \tau^d, x) Z_t (\mathrm{d}\tau^b, \mathrm{d}\tau^d , \mathrm{d}x) &= \sum_{k=1}^{N_t} f(I_k).
\end{aligned}
$$ {#eq-popZ}

The number of individuals present in the population *before time* $t$ is obtained by taking $f\equiv 1$:
$$
    N_t =  \int_{\mathcal{I}}  Z_t(\mathrm{d}\tau^b, \mathrm{d}\tau^d, \mathrm{d}x) = \sum_{k=1}^{N_t} \boldsymbol{1}_{\mathcal{I}} (I_k).
$$
Note that $(N_t)_{t\geq 0}$ is an increasing process since dead/exited individuals are kept in the population $Z$. The number of alive individuals in the population at time $t$ is: 
$$
    N_t^a =  \int_{\mathcal{I}}  \mathsf{1}_{\{\tau^d > t \} }Z_t(\mathrm{d}\tau^b, \mathrm{d}\tau^d, \mathrm{d}x) = \sum_{k=1}^{N_t} \mathsf{1}_{\{\tau^d_k > t \} }.
$$ {#eq-Nta}
Another example is the number of alive individuals of age over $a$ is
$$
    N_t([a,+\infty)) :=  \int_{\mathcal{I}}  \boldsymbol{1}_{[a,+\infty)}(t-\tau^b)\mathsf{1}_{]t,\infty]}(\tau^d) Z_t(\mathrm{d}\tau^b, \mathrm{d}\tau^d, \mathrm{d}x) = \sum_{k=1}^{N_t}  \boldsymbol{1}_{\{ t -\tau_k^b \geq a \}}\mathsf{1}_{\{\tau^d_k \geq t \} }.
$$

## Events {#sec-events}

The population composition changes at random dates following different types of events. `IBMPopSim` allows the simulation of IBMs with the following events types:

- A **birth** event at time $t$ is the addition of a new individual $I'=(t,\infty, X)$ of age $0$ to the population. Their date of birth is $\tau^b =t$, and characteristics is $X$, a random variable of distribution defined by the birth kernel $k^b(t,I,\mathrm{d}x)$ on ${\mathcal{X}}$, depending on $t$ and its parent $I$. The population size becomes $N_t = N_{t^-} + 1$, and the population composition after the event is $$Z_t  = Z_{t^-} +  \delta_{(t,\infty, X)}.$$

- An **entry** event at time $t$ is also the addition of an individual $I'$ in the population. However, this individual is not of age $0$. The date of birth and characteristics of the new individual $I'= (\tau^b, \infty, X)$ are random variables of probability distribution defined by the entry kernel $k^{en}(t, \mathrm{d}s, \mathrm{d}x)$ on ${\mathbb{R}}\times {\mathcal{X}}$. The population size becomes $N_t = N_{t^-} + 1$, and the population composition after the event is: 
$$
    Z_t  = Z_{t^-} +  \delta_{(\tau^b, \infty, X)}.
$$

- A **death** or **exit** event of an individual $I= (\tau^b,\infty, x)\in Z_{t^-}$ at time $t$ is the modification of its death date $\tau^d$ from $+\infty$ to $t$. This event results in the simultaneous addition of the individual $(\tau^b,t,x)$ and removal of the individual $I$ from the population. The population size is not modified, and the population composition after the event is 
$$
    Z_t  = Z_{t^-} +\delta_{(\tau^b,t,x)}- \delta_{I}.
$$

- A **swap** event (change of characteristics) results in the simultaneous addition and removal of an individual. If an individual $I= (\tau^b,\infty, x) \in Z_{t^-}$ changes of characteristics at time $t$, then it is removed from the population and replaced by $I' = (\tau^b,\infty, X)$. The new characteristics $X$ is a random variable of distribution $k^s(t, I,\mathrm{d}x)$ on ${\mathcal{X}}$, depending on time, the individual's age and previous characteristics $x$. In this case, the population size is not modified and the population becomes:
$$
    Z_t  = Z_{t^-}   +  \delta_{(\tau^b,  \infty, X)} -  \delta_{(\tau^b, \infty, x)}.
$$

To summarize, the space of event types is $E = \{ b, en, d, s \}$, and the jump $\Delta Z_t = Z_t - Z_{t^-}$ (change in the population composition) generated by an event of type $e \in \{ b, en, d, s \}$ is denoted by $\phi^e(t, I)$. We thus have the following rules summarized in the table @tbl-event-action.

Event | Type | $\phi^e(t, I)$ | New individual
------|:----:|----------------|---------------
Birth | $b$ | $\delta_{(t, \infty,  X)}$ | $\tau^b =t, \; X \sim k^b(t,I,\mathrm{d}x)$ 
Entry | $en$ | $\delta_{(\tau^b, \infty,  X)}$ | $(\tau^b, X) \sim k^{en}(t,\mathrm{d}s, \mathrm{d}x)$ 
Death/Exit | $d$ | $\delta_{(\tau^b, t,x)} - \delta_{(\tau^b, \infty, x)}$ | $\tau^d = t$
Swap | $s$ | $\delta_{(\tau^b, \infty , X)} - \delta_{(\tau^b, \infty, x)}$ | $X \sim k^s(t,I,\mathrm{d}x)$
: Action in the population for a given event name {#tbl-event-action}

::: {#rem-popfinale}

## Composition of the population

- At time $T$, the population $Z_T$ contains all individuals who lived in the population before $T$, including dead/exited individuals. If there are no swap events, or entries,the population state $Z_t$ for any time $t\leq T$ can be obtained from $Z_T$. Indeed, if $Z_T = \sum_{k=1}^{N_T}  \delta_{I_k}$, then the population at time $t\leq T$ is simply composed of the individuals born before $t$: $$
    Z_t = \sum_{k=1}^{N_T} \boldsymbol{1}_{\{\tau^b_k \leq t \}} \delta_{I_k}.
$$
- In the presence of entries (open population), a characteristic $x$ can track the individuals' entry dates. Then, the previous equation can be easily modified in order to obtain the population $Z_t$ at time $t\leq T$ from $Z_T$.

:::

## Events intensity {#sec-event-intensity}

## Pathwise representation
## Thinning of Poisson measure {#sec-thinning}
